---
title: "Lecture 1/8/2025"
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Elementary R

Examples of creating scalar variables and combining them:

```{r}
x=5
y=3
z = x+y
print(z)
```

```{r}
z
```

Strings are also possible:

```{r}
data_name = "day one example"
print(data_name)
```

`=` is equivalent to `<-` in R

```{r}
x<- 5
print(x)
```

```{r}
6->y
print(y)
```

# Arrays

Arrays are constructed using `c` for *combine*:

```{r}
data = c(1,2,3)
print(data)
```

```{r}
data_names = c("age", "height", "weight")
print(data_names)
```

Access elements with brackets with indexing beginning at 1 (not 0):

```{r}
data[1]
```

```{r}
data_names[2]
```

Whole array (or vectorized) operations are possible with numeric data:

```{r}
x=c(1., 2., 3.)
y=c(4., -2.4, 3.8)
```

```{r}
x+y
```

```{r}
x*y
```

Arrays of uniformly spaced elements:

```{r}
seq(0,1)
```

```{r}
seq(0,1,length=11)
```

eleven uniformly spaced points from 0 to 1, inclusive

```{r}
x = seq(0,1,length=11)
```

```{r}
y = sin(x)
y
```

R has, built in, classical functions like `exp`, `cos`, `log`, etc.

```{r}
print(log(10))
print(log10(10))
```

Exponentiation

```{r}
2^2
```

# Generating Random Variables in R

## Uniform Random Variables

These will be U(0,1) random variables (or more elegantly $U(0,1)$)

```{r}
runif(1)
```

This generates a single $U(0,1)$ random variable.

```{r}
runif(10)
```

```{r}
runif(50)
```

The index, `[k]` indicate the immediate next value is the k-th element in the array.

These examples are $U(0,1)$ examples. For $U(a,b)$, set the `min` and `max` values as desired:

```{r}
runif(1,min=-1, max=2) # U(-1,2) variable
```

```{r}
runif(5,min=-1, max=2) 
```

## Reproducibility and Seeds

```{r}
runif(1)
```

```{r}
runif(1)
```

Note that these are *pseudo*-random number generators. There is a deterministic algorithm lurking inside.

Reproducibility is achieved by setting the *seed*:

```{r}
set.seed(100)
```

```{r}
runif(5)
```

Reset the seed to the same value and call again:

```{r}
set.seed(100)
```

```{r}
runif(1)
```

```{r}
runif(2)
```

Changing the seed gives a difference sequence:

```{r}
set.seed(200)
```

```{r}
runif(1)
```

Set the seed at the beginning of any code using random number generation for reproducibility. Also essential for debugging and software development.

## Normal Random Variables

For $N(0,1)$:

```{r}
rnorm(1)
```

```{r}
rnorm(10)
```

In R, normal $N(m,\sigma^2)$ is parametrized by $m$ and $\sigma$, the standard dev:

```{r}
rnorm(5, 1, 0.1)
```

The above corresponds to $\sigma = 0.1$ not $\sigma^2 = 0.1$

# Markdown

Greek letters rendered in LaTeX:

$\sigma$, $\alpha$, $\beta$, $\gamma$ $\delta$, $\Sigma$, $\Xi$ , $\Psi$ $\psi$

and

$\sum$ and $\int$

For bits of code, use single quotes: `min` ,`f` , `mean`

Formatting and headings

*italics*

**bold**

## heading 2

Lists are formatted with `*`

-   First element

-   Second element

-   Third element

    -   tab in to create a sublist

Numbered lists are formatted with `1.`:

1.  First entry
2.  Second entry

# Elementary Statistics in R

Generate a basic numeric data set:

```{r}
set.seed(123)
xdata = rnorm(100)
print(xdata)
```

Basic summary statistics of data:

```{r}
summary(xdata)
```

More specific quantities:

```{r}
print(mean(xdata))
```

```{r}
print(median(xdata))
```

For variance:

```{r}
print(var(xdata))
```

For standard deviation:

```{r}
print(sd(xdata))
```

```{r}
print(quantile(xdata))
```

By default, `quantile` gives the 4 quartiles. Suppose we want the quintiles, instead, values at 0%, 20%, 40%, 60%, 80%, 100%:

1.  Create an array of values, `pvals` that contains the values at which we want the quantile function
2.  Evaluate `quantile(xdata, probs=pvals)`

```{r}
pvals = c(0, .2, .4, .6, .8, 1)
```

```{r}
quantile(xdata, probs = pvals)
```

```{r}
pvals2=c(0.2, .87, .99)
quantile(xdata, probs = pvals2)
```

# Control Structures in R

For loops, if-then-else's,

```{r}
for (j in 1:5){
  print(j)
}
```

```{r}
x = 5
if (x>3){
  print(x)
}else{
  print("nope")
}
```
