---
title: "Lecture 1/24/2025"
output: html_document
date: "2025-01-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ISLR2)
```

# Confidence Intervals

We will build CI's of the form (95% case)

$$
(\bar{x}_N - 1.96\sigma_N/\sqrt{N},\bar{x}_N + 1.96\sigma_N/\sqrt{N})
$$

or, more generally,

$$
(\bar{x}_N - z_{1-\alpha/2}\sigma_N/\sqrt{N},\bar{x}_N +  z_{1-\alpha/2}\sigma_N/\sqrt{N})
$$

and

$$
\text{Standard Error} = \frac{\sigma_N}{\sqrt{N}}
$$

where $\sigma_N$ is the sample standard deviation.

## Naive Computation

Generate some data, then compute the CI for the mean.

```{r}
set.seed(100)
x = rexp(100) # exponentially distributed random variables, not Gaussian
```

These are samples from $\mathrm{exp}(1)$, the exponential distribution with parameter $1$ by default. This has density $e^{-x}$ on $[0,\infty)$

```{r}
hist(x)
```

Pick an $\alpha$, $\alpha = 0.05$ and compute the CI (for the $1-\alpha$ % CI):

```{r}
alpha = .05
z = qnorm(1-alpha/2) # qnorm is the inverse CDF (quantile function) for N(0,1)
z
```

```{r}
qnorm(alpha/2) # NOTE, this the negative
```

The CI for the mean (really 1) is then:

```{r}
xbar = mean(x)
sigma= sqrt(var(x))
N = 100;
CI = c(xbar - z * sigma/sqrt(N),xbar + z * sigma/sqrt(N))
CI
```

CLAIM: True mean is 1 for this example

## T-Test

In the above computation, we assumed $N$ was large enough for the Central Limit Theorem to take over. But sometimes $N$ is too small, and/or the variance is too large. Use the t-test, or Student $t$ distribution to compensate. Like the Central Limit Theorem for small $N$

```{r}
t.test(x)
```

Defaults to 95% case (they all do).

```{r}
t.test(x,conf.level=0.99)
```

For $N$ large enough (typically $N>30$), T-Test will be the same as the CLT result.

# Data Frame Manipulation

Create a toy data frame:

```{r}
toy.df = tibble(x=1:3, y = 4:6)
```

```{r}
toy.df
```

Two features (columns), `x` and `y` , and 3 samples.

We want to add a new sample:

```{r}
add_row(toy.df, x=5, y = -8)
```

This did not alter `toy.df`. Need to assign it.

```{r}
toy.df
```

```{r}
toy2.df = add_row(toy.df, x=5, y = -8)
```

```{r}
toy.df
```

```{r}
toy2.df
```

Piping - language feature of R, and it's built around the command `|>` :

Example:

```{r}
toy3.df =toy.df |> add_row(x=5, y = -8)
toy3.df
```

Add a feature:

```{r}
toy.df |> add_column(day=c("Monday", "Friday", "Saturday"))
```

`toy.df` is unchanged:

```{r}
toy.df
```

Piping features and samples:

```{r}
toy.final.df = toy.df |> 
  add_column(day=c("Monday", "Friday", "Saturday")) |>  
  add_row(x=5, y = -8, day="Blaturday")
```

```{r}
toy.final.df
```

## Merging Data Frames

Working with the Wine data sets form [Wine Quality - UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/186/wine+quality). This has scores by wine experts and chemical measurements. Two challenges:

-   Two different data sets for red and white; read in each

-   Semi-colon separated, instead of comma separated; use `sep=";"`

```{r}
white.df = read.csv("~/code/Math_318_2025/data/wine+quality/winequality-white.csv", sep=";")  |>  
  as_tibble()
white.df
```

`read.csv` returns a default R data frame, we then pipe it in, `|>` to make a tibble.

```{r}
red.df = read.csv("~/code/Math_318_2025/data/wine+quality/winequality-red.csv", sep=";") |> 
  as_tibble()
red.df
```

Use `union_all` to combine data frames:

```{r}
attempt1.df = union_all(white.df, red.df)
attempt1.df
```

**NOTE** we have lost information as to wine color. Go back, add the feature and then join:

```{r}
white.df = white.df |>  add_column(color="white")
white.df
```

```{r}
red.df = red.df |> add_column(color="red")
red.df
```

Now combine:

```{r}
wine.df = union_all(white.df, red.df)
wine.df
```

## Cleaning Data

Create a data frame with a missing entry:

```{r}
x=1:10
y = x^2 
y[4]=NA # put a missing entry into the data
messy.df = tibble(x=x, y=y)
messy.df
```

Filter out samples with missing entries:

```{r}
clean.df = messy.df |>  na.omit() # drops samples with missing entries
clean.df
```

# More Data Frame Manipulation

Work with the `Auto` data set

```{r}
library(ISLR2)
```

```{r}
Auto |> as_tibble() -> Auto # converts to a tibble and saves it
```

```{r}
Auto |>filter(cylinders==3)
```

```{r}
Auto |>filter(cylinders==4)
```

```{r}
Auto |>filter(cylinders==3)|>select(year) # only return the specified feature
```

```{r}
Auto|>select(year)|>summary()
```

```{r}
Auto |>filter(cylinders==4)|>select(year) |> summary()
```

# Pivoting

Make the sample stock data:

```{r}
set.seed(100)
stocks.df = tibble(date=as.Date('2024-01-01')+0:9, "Apple"=100+cumsum(rnorm(10,0,4)),
                   "Google"=25+cumsum(rnorm(10,0,2)),
                   "Wawa"=10+cumsum(rnorm(10,0,0.5)))
```

```{r}
stocks.df
```

Closing prices of the three stocks on the specified dates.

Reasonable table format.

Easy to visualize individual stocks:

```{r}
plt = ggplot(stocks.df,mapping=aes(date,Apple))+geom_point()
print(plt)
```

But it would be nice to get all stocks at once, color coded, etc

## Pivot Longer

```{r}
stocks.long.df = stocks.df|>pivot_longer(cols=!date,
                        names_to="symbol",
                        values_to="price")
```

```{r}
stocks.long.df
```

`pivot_longer` replaces `gather`

```{r}
stocks.long.df = stocks.df|>gather(key=symbol, value=price,-date)

```

```{r}
stocks.long.df
```

```{r}
stocks.plt = ggplot(stocks.long.df,mapping=aes(date,price,color=symbol))+geom_point()
print(stocks.plt)
```

## Pivot Wider

Suppose we want to undo what we did

```{r}
stocks.long.df |> pivot_wider(names_from=symbol, values_from=price)
```

`pivot_wider` replaces `spread`

# Kables

To make nicely formatted tables that can be repsented, we use `kables` . These can be exported to Word, HTML, and Latex.

```{r}
library(knitr) # this provies the kable
library(kableExtra) # this provides formatting features for kables
```

```{r}
kable(stocks.df)
```

```{r}
kable(Auto) # this prints row "names", 1,2 ,3, 4...
```

```{r}
kable(Auto, row.names = FALSE)
```

Formatting kables:

```{r}
stocks.kb = kable(stocks.df, caption="Stocks Example")|> 
  kable_classic("striped", font_size=14)
stocks.kb
```

```{r}
stocks.kb = kable(stocks.df, caption="Stocks Example")|>  
  kable_classic(full_width=FALSE,"striped", font_size=14)
stocks.kb
```

```{r}
stocks.kb = kable(stocks.df, caption="Stocks Example")|> 
  kable_material(full_width=FALSE,"striped", font_size=10)
stocks.kb
```
