---
title: "Lecture 1/13/2025"
output: html_document
date: "2025-01-12"
---

Need to look at what this does.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse) # gets ggplot2 and tibble
```

# Better Graphics with ggplot2

## Visualizing the Auto Data Frame

The `Auto` data frame is part of the `ISLR2` package:

```{r}
library(ISLR2)
```

```{r}
Auto = as_tibble(Auto)
Auto
```

Before visualizing this, we will convert certain properties to categorical (factor) types:

```{r}
Auto$cylinders = as_factor(Auto$cylinders)
Auto$origin = as_factor(Auto$origin)
```

### Boxplots

This is a good data set to use boxplots on, as we can compare the fuel efficiency as a function of region of origin or number of cylinders in the engine:

```{r}
mpg.plt = ggplot(Auto) # we will reuse this 
mpg.plt1 = mpg.plt + geom_boxplot(mapping = aes(x=origin, y= mpg))
print(mpg.plt1)
```

```{r}
mpg.plt2 = mpg.plt + geom_boxplot(mapping = aes(x=cylinders, y= mpg))
print(mpg.plt2)
```

**Question** Why do 3 cylinder engines break the trend?

**NOTE** that we changed the horizontal grouping by changing the `x` argument to the `aes`. We can further slice our data, visualizing the breakdowns together:

```{r}
mpg.plt3 = mpg.plt + geom_boxplot(mapping = aes(x=cylinders, y= mpg, fill=origin))
print(mpg.plt3)

```

The `fill` argument further atomizes the `x` categories.

Within the 4 cylinder case, all regions are fairly similar in performance. it is the 8 cylinder engines from region 1 (North America) that weight down the average North American mpg performance.

```{r}
mpg.plt4 = mpg.plt + geom_boxplot(mapping = aes(x=origin, y= mpg, fill=cylinders))
print(mpg.plt4)

```

What's interesting about this plot is that you see that certain geographic regions only produce cars with some cylinder counts; only 4 and 6 cylinder engines are produced in all three regions.

### Histograms

We can perform similar analysis with histograms

```{r}
hist.mpg.plt1 = mpg.plt + geom_histogram(aes(x=mpg), bins=10) + ggtitle("MPG Distribution")
print(hist.mpg.plt1)
```

**NOTE:** I manually set `bins=10` , as in 10 histogram bins, because a warning was generated with the default value

```{r}
hist.mpg.plt2 = mpg.plt + geom_histogram(aes(x=mpg,fill=origin), bins=10) + 
  ggtitle("MPG Distribution by Origin")
print(hist.mpg.plt2)
```

The `fill` command atomizes the data in visualization (again).

Alternatively, we can set `color` :

```{r}
hist.mpg.plt3 = mpg.plt + geom_histogram(aes(x=mpg,color=origin), bins=10) + 
  ggtitle("MPG Distribution by Origin")
print(hist.mpg.plt3)
```

And we can improve the readability by making the bars transparent (changing `alpha` ):

```{r}
hist.mpg.plt4 = mpg.plt + geom_histogram(aes(x=mpg,fill=origin), bins=10, alpha=0.7) +
  ggtitle("MPG Distribution by Origin")
print(hist.mpg.plt4)
```

# Basic QQ Plots

Generate some data for comparison. Five data sets are generated in the following cell.

```{r}
set.seed(100)
u = runif(10^3) # U(0,1)
v = runif(2000, min=10, max=25) # U(10,25)
w = runif(100) # U(0,1)
x = rnorm(10^4) # N(0,1)
y = rnorm(5000, mean=10, sd =2) #N(10,2^2)
```

Use `qqplot` to compare data sets:

```{r}
qqplot(u,v)
```

This is a line, but it is not $y=x$, so these are not the same distribution.

```{r}
qqplot(u,w)
```

This is close to $y=x$, and it should be

```{r}
qqplot(u,x)
```

Comparison of $N(0,1)$ with $U(0,1)$; this is really not a line now. Unambiguously nonlinear

```{r}
qqplot(x,y)
```

Two different normals, they sit on a line, but it's not $y=x$

```{r}
qqnorm(x)
```

`qqnorm` checks a single data set against $N(0,1)$
