---
title: "Lecture 1/10/2025"
output:
  word_document: default
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

# Intro to Data Frames (tibbles)

These are built in, but we're going to use `tibble`, part of the `tidyverse` . If you do not have this installed, you can use `install.packages("tidyverse")` in the console

```{r}
library(tidyverse) # loads the entire package fore use, a la import from python
```

Create a data frame:

```{r}
cast=tibble("Names"=c("Jerry", "George", "Elaine", "Cosmo"), "Ages"=c(38, 38, 37, 999))
```

```{r}
cast
```

`<dbl>` is for double, as in double precision, the default data type for a real number in almost all modern languages.

`999` is a "garbage" value that really reflects that the measurement was unavailable. Try to clean out garbage values. Use `NA` (not available) value. Plays an analogous role to `Inf` or `NaN`

```{r}
cast$Names # access columns
```

```{r}
cast$Ages # access columns
```

Fix the entry that we believe is wrong:

```{r}
cast$Ages[4]=NA
```

```{r}
cast
```

Age should really be in an integer. Let's make sure knows that:

```{r}
cast$Ages = as.integer(cast$Ages) # this maps the doubles into integers and changes the data type
```

```{r}
cast
```

Data type is as desired.

```{r}
summary(cast)
```

This is an elementary data cleaning example.

# Read an Excel Sheet into R

The typcial work flow is read in a data set and load it into a `tibble`.

This builds on a `tidyverse` function

```{r}
library(readxl) # this provides Excel reading functions
```

```{r}
grades = read_excel("~/code/Math_318_2025/data/Grades.xlsx")
```

```{r}
grades
```

`<dbl>` refers to double precision floating point numbers (i.e., the reals).

```{r}
summary(grades)
```

We will return to how to get R to interpret different columns as different data types (i.e., categorical).

```{r}
grades[1,] # first row
```

```{r}
grades[15,] # 15th row
```

```{r}
grades[2] # with no comma, second column
```

## Setting Data Types in Data Frames

```{r}
grades
```

Note that both columns are stored as `<dbl>` , this is the double precision data type from C/C++/Java/Python/MATLAB/etc.

```{r}
summary(grades)
```

**NOTE** the statistics on the Student (ID), are meaningless.

We will let R know that it is a categorical variable:

```{r}
grades$Student = factor(grades$Student)
```

The above converts the data type of the column to `<fctr>` type, which is for categorical variables.

```{r}
grades
```

We can also set the Grade type to be an integer:

```{r}
grades$Grade=as.integer(grades$Grade)
```

```{r}
grades
```

Now summarize:

```{r}
summary(grades)
```

## Repeat this with CSV type data

CSV is more common/general than Excel.

```{r}
grades.csv = read.csv("../data/Grades.csv")
```

```{r}
grades.csv
```

This has been read in as a `df`, the default R, data frame (**NOT** yet a tibble). Usually worth converting to a tibble:

```{r}
grades.csv=as_tibble(grades.csv) # converts format
```

```{r}
grades.csv
```

Fix the data type of Student:

```{r}
grades.csv$Student=as_factor(grades$Student)
```

`as.factor` and `as_factor` are the same thing. True of many commands in R, `as.integer` and `as_integer` .

## Exporting data

```{r}
x=seq(0,1,length=11) # create some data
```

```{r}
x
```

```{r}
write.csv(x,file="../data/sample.csv") # writes x, and saves it to indicated file
```

Export summary data.

```{r}
grades.summary = summary(grades)
```

```{r}
grades.summary
```

```{r}
write.csv(grades.summary,file="~/code/Math_318_2025/data/grades.summary.txt") # writes only the summary information
```

How to export to LaTeX as a table

```{r}
library(xtable)
```

**NOTE** `xtable` expects a data frame type as its input argument. This can be a tibble or a default data frame. We must first map it to a data frame with `as.data.frame`

```{r}
xtable(as.data.frame(x))
```

Get into a text file:

```{r}
print.xtable(xtable(as.data.frame(x)), file="~/code/Math_318_2025/data/tex_table.txt")
```

# Basic Plotting and Visualization

## Built in functions

Make some basic data

```{r}
x=seq(0,1,length=51)
y = x*(1-x) # this is a parabola, x vs. y
```

Plot the parabola

```{r}
plot(x,y)
```

**NOTE** The default plotting is too bland for anything but a quick visualization.

Add some color and annotations:

```{r}
plot(x,y, main="Example Plot", xlab="x", ylab="x(1-x)", type="l", col="blue", lwd=2)
```

The additional arguments:

-   `main` - plot title

-   `xlab` - x axis label

-   `ylab` - y axis label

-   `type` - line with `l`, instead. of a scatter plot (default)

-   `col` - color

-   `lwd` - linewidth in pts.

What are our color options?

```{r}
colors()
```

```{r}
plot(x,y, main="Example Plot", xlab="x", ylab="x(1-x)", type="l", col="mediumorchid1", lwd=2)
```

Suppose we want to just export the figure:

```{r}
pdf(file="plot1.pdf", height=4, width=6) # note, need to specify height/width (in inches)
plot(x,y, main="Example Plot", xlab="x", ylab="x(1-x)", type="l", col="mediumorchid1", lwd=2)
dev.off() # stop writing to the file with this command
```

If the pdf format is ever too big, use `png` instead.

## Boxplots

```{r}
set.seed(100)
x=rnorm(100, mean=10, sd=2) # generate some data
```

```{r}
boxplot(x)
```

-   Solid line is the median

-   The box is the interquartile range, middle 50% of the data

-   Stems are the max/mins

-   If they show up, scatter points are "outliers"

Make this more interesting:

```{r}
boxplot(x, main="Sample Box Plot", horizontal = TRUE, 
        col="red", xlab="x", ylab="Samples 1", cex.lab=1.5, cex.axis=1.25, cex.main=2)
```

-   `horizontal` changes the orientation

-   `cex.lab` , `cex.axis` and `cex.main` - control the font sizes as percentages of the default.

## Histogram

```{r}
x=rnorm(10^4, mean=2, sd=2)
```

```{r}
hist(x)
```

Bell shaped Gaussian curve, as expected.

Annotate and improve:

```{r}
hist(x, breaks=50, main="N(2,2^2) Data", xlab="x", col="snow2", prob=TRUE)
```

By setting `prob=TRUE` , we are estimating the density, as in the probability density, from the data. Compare against analytic solution:

```{r}
hist(x, breaks=50, main="N(2,2^2) Data", xlab="x", col="snow2", prob=TRUE)
curve(dnorm(x, mean=2, sd=2), add=TRUE, col="navy", lwd=2)
```

`curve` plots a curve, `dnorm` specifies the function is the density of the normal, withe specified `mean` and `sd` , `add=TRUE` adds it to the existing plot.

Empirical histogram, when normalized, sits on the density curve.

# Better Graphics with ggplot2

As noted, the default graphics package in R is ok, but not great. Substantially better, out of the box, graphics can be generated with the `ggplot2` , which is part of the `tidyverse` :

```{r}
library(tidyverse)
```

The way of thinking of `ggplot2` graphics is in terms of addition, in that you will add graphical elements together, with the `+` , incrementally improving your plot. Roughly, you should think about it in terms of

Figure = Data Set + Graphical Manipulations of the data + Annotations

While it is not essential to create a data set, as a data frame (tibble or otherwise), to use `ggplot2` , there are advantages to using data frames here too.

## Parabola Plot

First, we will plot the parabola, $x(1-x)$ , manually:

```{r}
x = seq(0,1,length=51)
y = x*(1-x)
xy.plt = ggplot(mapping = aes(x,y))
```

This has created a plot, `xy.plt` which has, as its data, the *aesthetic*, `aes` . At this point, we see nothing. We can get the figure on the screen with the `print` command:

```{r}
print(xy.plt)
```

We now have a figure, but nothing in it. This is because we have not told it how to, geometrically, interpret the data.

### Scatter Plot

We have several options:

`geom_point()` for a scatter plot.

```{r}
xy.plt.point = xy.plt + geom_point() # interpret the data with a scatter plot
print(xy.plt.point)
```

### Line Plot

```{r}
xy.plt.line = xy.plt + geom_line() # interpret the data with a line plot
print(xy.plt.line)
```

### Both

```{r}
xy.plt.scatter_line = xy.plt +geom_point()+ geom_line() # interpret the data with a line plot
print(xy.plt.scatter_line)
```

### Titles

We can add a title to the plot with `ggtitle` :

```{r}
xy.plt.title = xy.plt + geom_line() + ggtitle("Parabola Plot") 
# interpret the data with a line plot
print(xy.plt.title)
```

### Saving to Disk

Figures can be saved to disk with the `ggsave` command:

```{r}
ggsave("ggplt1.pdf", plot = xy.plt.title, width = 6, height = 4) # size in inches
```

### Working with Data Frames

As noted, `ggplot2` plays nicely with data frames. First, we will stick our data into a data frame and then build on that:

```{r}
xy.df=tibble(Position=x,Velocity=y) # These are sample variable names
xy.df
```

```{r}
df.plt = ggplot(xy.df) + # load data
  geom_line(mapping=aes(x=Position,y=Velocity), color="darkorange1",lwd=2) + # specify geometry
  ggtitle("Plotting with a data frame")
print(df.plt)
```

Notice that it automatically picks up the variable names from the data frame for the axes

## Grades Data Set example

```{r}
hist_plt = ggplot(data=grades)
hist_plt = hist_plt + geom_histogram(mapping = aes(x=Grade),bins = 10)
hist_plt = hist_plt + ggtitle("Grade Distribution")
print(hist_plt)
```

## 
