---
title: "Lecture 1/15/2025"
output: html_document
date: "2025-01-13"
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ISLR2)
```

# Basic QQ Plots

```{r}
set.seed(100)
u = runif(10^3) # U(0,1)
v = runif(2000, min=10, max=25) # U(10,25)
w = runif(100) # U(0,1)
x = rnorm(10^4) # N(0,1)
y = rnorm(5000, mean=10, sd =2) #N(10,2^2)
```

## QQ Plots with ggplot2

### Single Data Set

Recall, this compares against $N(0,1)$

```{r}
qq.plt = ggplot(mapping=aes(sample=u)) + 
  geom_qq() + geom_qq_line(color="red") + 
  ggtitle("QQ Plot for a U(0,1) Data")
print(qq.plt)
```

```{r}
qq.plt = ggplot(mapping=aes(sample=y)) + 
  geom_qq() + geom_qq_line(color="red") + 
  ggtitle("QQ Plot for Gaussian Data")
print(qq.plt)
```

This sits on the $y=2x+10$ line, since the data is from $N(10,2^2)$

### Two Data Sets

This is not built in to `ggplot2` and requires a bit of programming.
First, we `sort` the data:

```{r}
su = sort(u) # sort array in x_i<x_{i+1} ordering
sv = sort(v)
sx = sort(x)
```

Next, we will need to interpolate (using `approx`), if the lengths of
the arrays we wish to compare are not equal:

```{r}
if (length(su)< length(sv)){
  su_interp=su
  # perform a linear interpolation to get a vector of the same length
  sv_interp=approx(1:length(sv),sv,n=length(su))$y
}else{
  sv_interp=sv
  su_interp=approx(1:length(su),su,n=length(sv))$y
}
```

For conciseness, pack into a `tibble`:

```{r}
uv.df <- tibble(u=su_interp, v=sv_interp)
```

For comparing `u` against `v`

Now plot:

```{r}
qq2.plt =ggplot(data=uv.df) + 
  geom_point(aes(x=u, y=v)) + 
  ggtitle("QQ Plot of Two Distinct Uniform Data Sets")
print(qq2.plt)

```

Repeat with `u` against `x`:

```{r}
if (length(su)< length(sx)){
  su_interp <- su
  # perform a linear interpolation to get a vector of the same length
  sx_interp <- approx(1:length(sx),sx,n=length(su))$y
}else{
  sx_interp <- sx
  su_interp <- approx(1:length(su),su,n=length(sx))$y
}
ux.df <- tibble(u=su_interp, x=sx_interp)

```

Plot:

```{r}
qq2ux.plt =ggplot(data=ux.df) + 
  geom_point(aes(x=u, y=x)) + 
  ggtitle("QQ Plot of Normal and Uniform Data")
print(qq2ux.plt)

```

# Additional Annotation Tools

Manipulate the `Auto` data set

```{r}
Auto$origin=as_factor(Auto$origin)
Auto$cylinders=as_factor(Auto$cylinders)
```

```{r}
mpg.plt = ggplot(Auto) # we will reuse this 
mpg.plt1 = mpg.plt + geom_boxplot(mapping = aes(x=origin, y= mpg))
print(mpg.plt1)
```

```{r}
summary(Auto$origin)
```

This could be in a caption or the text, but can we get into the figure
directly?

Add an annotation:

```{r}
mpg.plt2 = mpg.plt1 + annotate("text", x=c(1.1, 2.1, 3.1), y = c(30, 40, 45), 
                               label=sprintf("n=%d", summary(Auto$origin)),
                               color="red")
print(mpg.plt2)
```

What did we do:

-   `text` - says the figure annotation will be a text object

-   `x` and `y` - these are selected by guess and check as to where to
    put the labels

-   `label=sprintf` - this is C/C++ style string formatting.

Alternatively, turn the summary information into a data frame itself:

```{r}
group_by(Auto, origin)
```

```{r}
tally(group_by(Auto, origin))
```

```{r}
summary.df= tally(group_by(Auto, origin))
summary.df
```

```{r}
tally(Auto)
```

```{r}
Auto %>% group_by(cylinders) %>%tally()
```

Alternatively, uses pipes: `%>%`

```{r}
summary.df = Auto %>% group_by(origin) %>% tally()
```

```{r}
summary.df
```

`tally` sums up the data in each case from the `group_by`.

```{r}
auto.plt = ggplot(Auto) +
  geom_boxplot(aes(x=origin, y = mpg)) + 
  geom_label(data=summary.df, aes(x=origin, label=sprintf("n = %d",n)), y =max(Auto$mpg), color="red")

print(auto.plt)
```
