---
title: "Lecture 2/26/2025"
output: html_document
date: "2025-02-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(boot)
```

# Bootstrap

Application to estimating the kurtosis of a random variable:

$$
\text{Kurtosis} = \frac{\mathbb{E}[(X-\mu)^4]}{\mathbb{E}[(X-\mu)^2]^2}
$$

This measures how far from normal a distribution is. A naive estimate is:

$$
\frac{\frac{1}{n}\sum_{i=1}^n (x_i-\hat\mu)^4}{(\frac{1}{n}\sum_{i=1}^n (x_i-\hat\mu)^2)^2}
$$

An error bound on our estimate of this can be formed by bootstrap methods.

## Generate data

We will estimate the kurtosis of the $\mathrm{Exp}(1)$, the exponential distribution with parameter $\lambda = 1$

First, we generate data for this experiment

```{r}
set.seed(100)

# number of samples in the data set
n <- 10^3
x <- rexp(n) # these are exponential random variables (non-Gaussians)

x.df <- tibble(x=x)
```

True kurtosis of this problem is 9

## Define Function for Bootstrapping

```{r}
kurtosis.fn = function(data,index){
  # extract values of interest in the current bootstrap sample determined by index
  x = data$x[index]
  # estimate the mean
  mu = mean(x)
  # compute the estiamted kurtosis
  k = mean((x-mu)^4)/(mean((x-mu)^2)^2)
  return (k)
}
```

## Compute Bootstrap

```{r}
bt = boot(x.df,kurtosis.fn,R=10^3) # R is the number of Bootstrap samples

```

```{r}
bt
```

```{r}
boot.ci(bt)
```

4 different types of CIs for the estimated value. True answer, 9, is in all of them.
