---
title: "Lecture 1/29/2025"
output: html_document
date: "2025-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ISLR2)
```

# Linear Regression

Generate data:

```{r}
set.seed(100) # for random noise
x= seq(0, 10, length=15) # 15 uniformly spaced points between 0 and 10
beta0 = 2.
beta1 = -1.5
y = beta0 + beta1 * x + 2 * rnorm(length(x)) # Y = β0 + β1 * X + ε
df = tibble(x=x, y=y)
df
```

```{r}
lm1 = lm(y~x, df)
```

## Assessing goodness of fit

```{r}
summary(lm1)
```

# Confidence and Prediction Intervals for the Model

## Confidence Intervals

```{r}
confint(lm1)
```

These are fairly broad for the true value of $\beta_0$ and $\beta_1$.

The shaded region in this figure defaults to 95% CI for $\beta_0 +\beta_1 X$

```{r}
plt = ggplot(df, aes(x=x,y=y)) + geom_point()+geom_smooth(method=lm)
print(plt)
```

## Prediction Intervals

Construct a prediction interval:

```{r}
predict(lm1,df, interval="prediction")
# predict(lm1,df, interval="prediction", level = .99)
```

Each row corresponds to the $x_i$ in the df , giving the estimated $\hat{\beta_0} + \hat{\beta_1}x_i$ and the upper and lower prediction interval (defaulting to 95%) on the corresponding $y_i$.

## Visualization

To get into the figure, store in a data frame:

```{r}
pred.df = as_tibble(predict(lm1,df, interval="prediction"))
```

```{r}
plt = ggplot(df, aes(x=x,y=y)) + geom_point()+geom_smooth(method=lm) + 
  geom_ribbon(data = pred.df, aes(ymin = lwr, ymax = upr), fill="red", alpha = .2)
print(plt)
```

alpha is the transparency level.

CI is measuring the uncertainty in the trend line, $\hat{\beta}_0 +\hat{\beta}_1 x$ and the prediction interval measure the uncertainty in $y$

Alternatively way of capturing the prediction interval (with lines):

```{r}
plt = ggplot(df, aes(x=x,y=y)) + geom_point()+geom_smooth(method=lm) + 
  geom_line(data = pred.df, aes(y=lwr),linetype="dashed") +
  geom_line(data = pred.df, aes(y=upr),linetype="dashed")
print(plt)
```
