---
title: "Lecture 1/31/2025"
output: html_document
date: "2025-01-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ISLR2)
```

# Multilinear Regression

## Elementary Example

Learn

$$
Y = 5 + 2X_1 -3X_2 + \epsilon
$$

$p=2$ and we have noise in our measurements.

Create synthetic data:

```{r}
n = 25
beta0 = 5
beta1 = 2 
beta2 = -3

set.seed(123)
# create random points, these are arbitrary, we could use normals, or just seq
x1 = runif(n, min=-2,2)
x2 = runif(n, min=-2,2)

y = beta0 + beta1 * x1 + beta2 * x2 + rnorm(n, sd = 0.5)
```

Need to move this into a data frame

```{r}
example1.df = tibble(x1=x1, x2 =x2, y = y)
example1.df
```

Perform the regression:

```{r}
example1.lm = lm(y~x1 + x2,example1.df) # note implicitly includes the constant
```

NOTE: the only thing we needed to do to make this multilinear was to put in the formula `y~x1+x2` , with two variables.

```{r}
example1.lm
```

Check the quality of the fit:

```{r}
summary(example1.lm)
```

Note: The $R^2 \approx .99$, so this is a very good fit. All three coefficients are statistically significant.

```{r}
confint(example1.lm)
```

The 95% confidence intervals of the coefficients all include the truth.
